\label{sec:adders}
Every computer is an electric device which changes the state of its memory (i.e., randomly-accessible memory (RAM) or on-chip memory also referred to as {\em registers}) in response to the execution of instructions provided to the computer through RAM. The change of the state is performed by a unit called the {\em arithmetic logic unit} (ALU) which implements at the very least $n$-bit addition, bit-wise logic operations (i.e., AND, OR, NOT) as well as bit-wise shift operations\footnote{Computer architecture differs in the range of instructions is supports in hardware---the so-called instruction set architecture (ISA)---as well as the way that some operations are supported on RAM or only on registers.}. Note that the instructions themselves are part of the state of the computer and that the program flow through this state is handled by using the ALU on a special register called the {\em program counter} (always pointing to the location in RAM where the current instruction is contained). Thus, the ALU is the most central component that allows any computer to perform all its tasks.

\begin{figure}
    \begin{minipage}[c]{.65\linewidth}
        \centering
        \begin{tikzpicture}
            % Circuit style
            \ctikzset{
                logic ports=ieee,
                logic ports/scale=0.8,
                % logic ports/fill=lightgray
            }

            % Logic ports
            \node[nor port] (NOR) at (0,0){};
            \node[nand port] (NANDa) at (0,-2){};
            \node[not port] (NOTa) at (2,0){};
            \node[not port] (NOTb) at (2,-2.6){};
            \node[not port] (NOTc) at (6,-1){};
            \node[nand port] (NANDb) at (4,-1){};

            % Input and output ports
            \node (a) at (-2,-0.6) [left] {$a$};
            \node (b) at (-2,-1.2) [left] {$b$};
            % \node (c1) at (1,-3) [above] {$\cout$};
            \node (d) at (1,0) [above] {$d$};
            \node (e) at (1,-2) [above] {$e$};
            \node (f) at (2.8,0) [above] {$f$};
            \node (g) at (5.1,-1) [above] {$g$};
            \node (s) at (7,-1) [right] {$s$};
            \node (c2) at (7,-2.6) [right] {$\cout$};
            \node (af) [right = 0.5 of a, coordinate] [left] {};
            \node (bf) [right = 1 of b, coordinate] [left] {};

            % % Connection
            \draw (NANDa.out) -| (NOTb.in);
            \draw (NOR.out) -- (NOTa.in);
            \draw (NOTa.out) -| (NANDb.in 1);
            \draw (NANDa.out) -| (NANDb.in 2);
            \draw (NANDb.out) -| (NOTc.in);
            \draw (NOTc.out) -* (s);
            \draw (NOTb.out) -* (c2);

            \draw (NOR.in 1) -| (af) |- (NANDa.in 2);
            \draw (NOR.in 2) -| (bf) |- (NANDa.in 1);
            \draw (a) -- (af);
            \draw (b) -- (bf);
            % \draw (c1) |- (c2);
        \end{tikzpicture}
    \end{minipage}
    \hfill
    \begin{minipage}[c]{.34\linewidth}
        \centering
        \begin{tabular}{c|c||c|c|c|c|c|c}
            $a$ & $b$ & $d$ & $e$ & $f$ & $g$ & $s$ & $\cout$ \\
            \hline
            $0$ & $0$ & $1$ & $1$ & $0$ & $1$ & $0$ & $0$     \\
            $0$ & $1$ & $0$ & $1$ & $1$ & $0$ & $1$ & $0$     \\
            $1$ & $0$ & $0$ & $1$ & $1$ & $0$ & $1$ & $0$     \\
            $1$ & $1$ & $0$ & $0$ & $1$ & $1$ & $0$ & $1$     \\
        \end{tabular}
    \end{minipage}
    \caption{{\bf (Left)} Logic circuit design of a half-adder function in terms of NAND, NOR and NOT gates that computes the sum of two binary numbers $a \in \Bin$ and $b \in \Bin$ where $s$ is the sum of $a$ and $b$ and $\cout$ captures the carry-bit indicating an overflow (i.e., if $a$ and $b$ are both $1$, then $s=0$ and $\cout=1$). {\bf (Right)} The logical equivalent of the half-adder with the values for all intermediate results $d \in \Bin$, $e \in \Bin$, $f \in \Bin$, and $g \in \Bin$ for every possible list of (binary) input values. Note that $s = a \oplus b$ and $\cout = a \land b$. \label{fig:half-adder}}
\end{figure}

We will focus on one operation of the ALU that is central to all operations namely {\em addition}. Depending on the native bandwidth of the CPU, this includes the addition of two $4$-, $8$-, $16$-, $32$- or even $64$-bit numbers provided in registers. As each number is represented by bits, we will denote the two inputs to addition $A=a_0a_1a_2a_3\cdots a_n$ and $B=b_0b_1b_2b_3\cdots b_n$, respectively. The addition of $A$ and $B$ is performed in a pipeline of $n$ steps:
\begin{enumerate}
    \item {\bf Half-Adder}. First, $a_0$ and $b_0$ get added according to the rules of binary addition resulting in a sum bit $s_0$ as well as a carry-out bit $\cout^0$.
    \item {\bf Full-Adder}. Second, $a_1$, $b_1$ and $\cout^0$ get added according to the rules of binary addition with a carry-in bit $\cin^1=\cout^0$ resulting in the sum bit $s_1$ as well as the carry-out bit $\cout^1$.
    \item {\bf Ripple-Carry Adder}. This sequence is repeated until the $n$th position given the bit string $s_0s_1s_2s_3\cdots s_n\cout^n$ which is the $(n+1)$-bit answer of the addition of $A$ and $B$.
\end{enumerate}


\begin{figure}
    \begin{minipage}[c]{.6\linewidth}
        \centering
        \begin{tikzpicture}
            % Circuit style
            \ctikzset{
                logic ports=ieee,
                logic ports/scale=0.8,
                % logic ports/fill=lightgray
            }

            % Logic ports
            \node[draw,minimum width=1.5cm,minimum height=2cm] (HA1) at (0,-1.21){HA$_1$}
            ($(HA1.west)!0.6!(HA1.north west)$) ++(-0.25,0) coordinate (HA1in1)
            ($(HA1.west)!0.6!(HA1.south west)$) ++(-0.25,0) coordinate (HA1in2)
            ($(HA1.east)!0.6!(HA1.north east)$) ++(0.25,0) coordinate (HA1out1)
            ($(HA1.east)!0.6!(HA1.south east)$) ++(0.25,0) coordinate (HA1out2);
            \node[draw,minimum width=1.5cm,minimum height=2cm] (HA2) at (2.2,0){HA$_2$}
            ($(HA2.west)!0.6!(HA2.north west)$) ++(-0.25,0) coordinate (HA2in1)
            ($(HA2.west)!0.6!(HA2.south west)$) ++(-0.25,0) coordinate (HA2in2)
            ($(HA2.east)!0.6!(HA2.north east)$) ++(0.25,0) coordinate (HA2out1)
            ($(HA2.east)!0.6!(HA2.south east)$) ++(0.25,0) coordinate (HA2out2);
            \node[nor port] (OR) at (4.5,-1.59){};
            \node[not port] (NOT) at (6.2,-1.59){};

            \draw (HA1in1) -- ++(0.25,0);
            \draw (HA1in2) -- ++(0.25,0);
            \draw (HA2in1) -- ++(0.25,0);
            \draw (HA2in2) -- ++(0.25,0);
            \draw (HA1out1) -- ++(-0.25,0);
            \draw (HA1out2) -- ++(-0.25,0);
            \draw (HA2out1) -- ++(-0.25,0);
            \draw (HA2out2) -- ++(-0.25,0);

            % % Input and output ports
            \node (a)  [left = 0.21 of HA1in1] [left] {$a$};
            \node (b)  [left = 0.21 of HA1in2] [left] {$b$};
            \node (cin)[left = 2.36 of HA2in1] [left] {$\cin$};
            \node (d) [left = 0.1 of HA2in2] [above] {$d$};
            \node (e) [left = 0 of OR.in 2] [below] {$e$};
            \node (f) [left = 0 of OR.in 1] [above] {$f$};
            \node (g) [left = 0 of NOT.in 1] [above] {$g$};
            \node (s) [right = 3.8 of HA2out1] {$s$};
            \node (c_out) [right = 0.1 of NOT.out] [right] {$\cout$};
            \node (af) [right = 0.5 of a, coordinate] [left] {};
            \node (bf) [right = 1 of b, coordinate] [left] {};

            % % % Connection
            \draw (a) -- (HA1in1);
            \draw (b) -- (HA1in2);
            \draw (HA1out1) -- (HA2in2);
            \draw (cin) -| (HA2in1);
            \draw (HA2out2) |- (OR.in 1);
            \draw (HA1out2) |- (OR.in 2);
            \draw (OR.out) -- (NOT.in);
            \draw (NOT.out) -- (c_out);
            \draw (HA2out1) -- (s);
        \end{tikzpicture}
    \end{minipage}
    \hfill
    \begin{minipage}[c]{.38\linewidth}
        \centering
        \begin{tabular}{c|c|c||c|c|c|c|c|c}
            $a$ & $b$ & $\cin$ & $d$ & $e$ & $f$ & $g$ & $s$ & $\cout$ \\
            \hline
            $0$ & $0$ & $0$    & $0$ & $0$ & $0$ & $1$ & $0$ & $0$     \\
            $0$ & $1$ & $0$    & $1$ & $0$ & $0$ & $1$ & $1$ & $0$     \\
            $1$ & $0$ & $0$    & $1$ & $0$ & $0$ & $1$ & $1$ & $0$     \\
            $1$ & $1$ & $0$    & $0$ & $1$ & $0$ & $0$ & $0$ & $1$     \\
            $0$ & $0$ & $1$    & $0$ & $0$ & $0$ & $1$ & $1$ & $0$     \\
            $0$ & $1$ & $1$    & $1$ & $0$ & $1$ & $0$ & $0$ & $1$     \\
            $1$ & $0$ & $1$    & $1$ & $0$ & $1$ & $0$ & $0$ & $1$     \\
            $1$ & $1$ & $1$    & $0$ & $1$ & $0$ & $0$ & $1$ & $1$     \\
        \end{tabular}
    \end{minipage}
    \caption{{\bf (Left)} Logic circuit design of a full-adder function that computes the sum of two binary numbers $a \in \Bin$ and $b \in \Bin$ when a carry-in bit $\cin \in \Bin$ is also passed on. {\bf (Right)} The logical equivalent of the full-adder with the values for all intermediate results $d \in \Bin$, $e \in \Bin$, $f \in \Bin$ and $g \in \Bin$ for every possible list of the eight (binary) input values. \label{fig:full-adder}}
\end{figure}

In Figure \ref{fig:half-adder} and \ref{fig:full-adder} we have shown both the circuit design as well as the logic table for a 1-bit half-adder and full-adder. In terms of logic circuits, the sum bit of the half-adder is nothing more than and XOR of both inputs $a$ and $b$ and the carry-out bit is the logical conjunction of the two inputs. For the full-adder, the sum bit is the XOR of the carry-in bit, $\cin$, and the XOR of both inputs $a$ and $b$; that is, if the carry-in bit is zero, the sum is the same as that of a half-adder but if the carry-in bit is one, the sum is the negation of the sum bit of the half-adder. Similarly, for the full-adder, the carry-out bit, $\cout$ is one if either both inputs $a$ and $b$ are one or if the carry-in bit, $\cin$ is one and at least one of $a$ and $b$ are one. More formally,
\begin{align}
    s & = a \oplus b               & \cout & = a \land b                                                                \\
    s & = (a \oplus b) \oplus \cin & \cout & = (a \land b) \lor ((a \lor b) \land \cin) \label{eq:full_adder_logic} \,,
\end{align}
where the first line is for the half-adder and the second line is for the full-adder, respectively.

Regarding the ripply-carry adder, looking at the rule for $\cout$ in (\ref{eq:full_adder_logic}) we can see that $(a \land b) = 1$ is {\em forcing} the carry-out bit $\cout$ to be one---irrespective of the value of $\cin$--- and that $(a \lor b) = 1$ is {\em propagating} the carry-out bit from left to right (i.e., if $\cout = \cin$). One can use these two properties when parallelizing additions over many bits because the individual conjunctions $a_i \land b_i$ do not need to wait for the carry-in bit from the prior positions (which would be required in the ripple-carry adder) and can be computed in parallel; similarly, the individual disjunctions $a_i \lor b_i$ can be computed ahead of time all that needs to happen in sequence is a correction of carry-out flags where the disjunctions were one. This circuit design is faster and is known as a {\em carry-lookahead} adder; however, we will not consider this design here.

Finally, it is worth noticing that we expressed the half-adder, full-adder and ripple-carry adder in terms of the basic logic circuits of NAND, NOR and NOT (rather than AND, OR and NOT). The reason for this choice is that we will present a realistic noise model of the adders for an {\em actual} implementation of the logic circuits on {\em complementary metal-oxide-semiconductor (CMOS)}, the most commonly used fabrication process which uses complementary and symmetrical pairs of p-type and n-type {\em metal-oxide-semiconductor field-effect transistors (MOSFETs)} for logic functions. In CMOS, the smallest circuit is the NOT-gate which needs exactly one one p-type and n-type MOSFET in series: if the input voltage is zero (i.e., shorted w.r.t.\ ground), the n-type MOSFET is blocking and the current flows from the source to the output and therefore the output is non-zero (i.e., not shorted w.r.t.\ ground). Similarly, if the input voltage is above the threshold (i.e., not shorted w.r.t.\ ground), the p-type MOSFET is blocking and the current flows from input to ground resulting in a voltage of zero at the output. Similarly, in CMOS, the next smallest (elementary) circuit are the NAND- and NOR-gate which require two p-type and two n-type MOSFETs either in series or parallel (but always in opposite ways for a single circuit). Every other logical function is a composition of these three elementary logic gates and we therefore present all our analysis of adders in terms of these basic logic gates\footnote{Note that in terms of Boolean algebra, NAND or NOR are the {\em smallest} set of logic functions that can express any Boolean expression because $\neg a = \neg(a \land a)$ but this type of analysis would not yield a realistic noise model and does not correspond to the mapping of a logic equation to CMOS.}.